Computing MEM seeding probabilities

### 2019-02-15 MEM.tex
Added some references about seeding.

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to some other location,
$iii.$ the DNA fragment is not mapped at all.

The most popular heuristic to address the best location problem is to
first identify short regions of high similiarity between the read and the
genome, and then run an exact alignment algorithm around those regions.
The first step is usually called ``seeding'' and the regions of high
similarity are called ``seeds''. This is for instance the general strategy
of the popular alignment algorithm BLAST\cite{pmid2231712}.

@@

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
one can develop faster mapping algorithms at no cost for accuracy.

@@

In general, seeds do not have to be exact matches\cite{pmid20460430}. For
instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding'' (where MEM stands for Maximal Exact Match). This approach gives
good empirical results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify MEM seeds, even when the reference
genome is very large\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are available
at all times. Looking ahead to section~\ref{sec:est}, let us just mention
that this can be done with an algorithm called the \emph{backward
search}\cite{ferragina2000opportunistic}.

@@

We turn to the question of translating the extended MEM segment
representation to a construction of reads without good seeds. For this, we
use tools borrowed from analytic combinatorics. We give the minimum amount
of background that is necessary to expose the theory. A more detailed
account of the strategy to compute seeding probabilities with analytic
combinatorics has been published
elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
readers interested in learning more about analytic combinatorics to the
textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.

@@

The number of errors $X$ in a read of size $k$ has a Binomial
distribution $X \sim B(k,p)$. From\cite{arratia1989tutorial} we can bound
the probabilities of the tail with the expression

@@

The method is summarized by the algorithm below. It requires efficient
methods to generate random samples with a geometric distribution and a
binomial distribution. Sampling from a geometric distribution can be done
by computing the logarithm of a uniform $(0,1)$ random variable. Sampling
from a binomial distribution can be done by the method of
Kachitvichyanukul and Schmeiser\cite{kachitvichyanukul1988binomial}.

@@

$P(E_0)$ is straightforward to evaluate because it involves a single
sequence. Methods to do this efficiently have been explained in detail
elsewhere\cite{filion2018analytic}. Here we only give the recurrence
equation to compute $P(E_0)$. Assume that the probability of a sequencing
error $p$ is known, that $q = 1-p$, and that $a_k$ is the probability that
a read of size $k$ does not contain a match of size $\geq \gamma$ for the
target. Then the following equation can be used to compute $a_k$ by
recurrence:

@@

The backward search is a method that was first proposed by Ferragina and
Mazini\cite{ferragina2000opportunistic} to identify exact matches between
a text query and an index. The query is extended one letter/nucleotide at
a time, and at each iteration the search returns the number of matches in
the index/genome. This is thus a very efficient method to identify MEM
seeds, because one can keep extending the query until there is no match in
the genome or until the end of the read.

@@

\bibliography{references,pubmed}
\bibliographystyle{plain}

>>>>

There exists exact algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location~\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to some other location,
$iii.$ the DNA fragment is not mapped at all.

The most popular heuristic for this problem is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm around those regions. The first step is a
filtration called ``seeding'' and the regions of high similarity
are called ``seeds''. This is for instance the general strategy of the
popular alignment algorithm BLAST~\cite{pmid2231712}.

@@

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others (see~\cite{sun2005designing,
pmid11934743,xu2006optimizing,pmid17044164,brejova2003vector,
pmid18684737} for examples of design, see~\cite{pmid16533404,pmid20460430}
for comparisons and discussions, and see~\cite{navarro2001guided} for a
more general review of filtration methods in pattern matching). The upshot
is that one can develop faster mapping algorithms at no cost for accuracy.

@@

In general, seeds do not have to be exact matches~\cite{pmid20460430}. For
instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding'' (where MEM stands for Maximal Exact Match). This approach gives
good empirical results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify MEM seeds, even when the reference
genome is very large~\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are available
at all times. Looking ahead to section~\ref{sec:est}, let us just mention
that this can be done with an algorithm called the \emph{backward
search}~\cite{ferragina2000opportunistic}.

@@

We turn to the question of translating the extended MEM segment
representation to a construction of reads without good seeds. For this, we
use tools borrowed from analytic combinatorics. We give the minimum amount
of background that is necessary to expose the theory. A more detailed
account of the strategy to compute seeding probabilities with analytic
combinatorics has been published
elsewhere~\cite{filion2017analytic,filion2018analytic}. We refer the
readers interested in learning more about analytic combinatorics to the
textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.

@@

The number of errors $X$ in a read of size $k$ has a Binomial
distribution $X \sim B(k,p)$. From~\cite{arratia1989tutorial} we can bound
the probabilities of the tail with the expression

@@

The method is summarized by the algorithm below. It requires efficient
methods to generate random samples with a geometric distribution and a
binomial distribution. Sampling from a geometric distribution can be done
by computing the logarithm of a uniform $(0,1)$ random variable. Sampling
from a binomial distribution can be done by the method of
Kachitvichyanukul and Schmeiser~\cite{kachitvichyanukul1988binomial}.

@@

$P(E_0)$ is straightforward to evaluate because it involves a single
sequence. Methods to do this efficiently have been explained in detail
elsewhere~\cite{filion2018analytic}. Here we only give the recurrence
equation to compute $P(E_0)$. Assume that the probability of a sequencing
error $p$ is known, that $q = 1-p$, and that $a_k$ is the probability that
a read of size $k$ does not contain a match of size $\geq \gamma$ for the
target. Then the following equation can be used to compute $a_k$ by
recurrence:

@@

The backward search is a method that was first proposed by Ferragina and
Mazini~\cite{ferragina2000opportunistic} to identify exact matches between
a text query and an index. The query is extended one letter/nucleotide at
a time, and at each iteration the search returns the number of matches in
the index/genome. This is thus a very efficient method to identify MEM
seeds, because one can keep extending the query until there is no match in
the genome or until the end of the read.

@@

\bibliography{references,pubmed}
\bibliographystyle{unsrt}
